Pragmatic interactions lead to efficient language structure and use

### 2019-01-25 paper/paper.Rmd
new draft for MF/NDG with partially complete supplementary materials

Why do languages look they way they do? Zipf (1935) proposed that distributional properties found in natural language were evidence of speaker-listener effort minimization. In his own words, “we are arguing that people do in fact act with a maximum economy of effort, and that therefore in the process of speaking-listening they will automatically minimize the expenditure of effort.” Evidence for this claim was largely derived at the level of the lexicon. Zipf argued that the particular relationship between a word’s frequency and its rank, length, and denotation size could be explained as an emergent property of speaker-listener effort minimization. \par

Zipf articulated what is now considered a *functionalist* approach to language science -- analyzing language structure and use in terms of efficiency. Such an approach might reframe our opening question as follows -- how does having \textbf{property x} make using language $\ell$ more or less useful for communication? This reframing provides an opportunity to study both language *structure* (as Zipf primarily did) or *use* in terms of efficiency. For example, Regier et al. (201X) showed that languages appear to organize semantic domains (form-meaning mappings) in maximally efficient ways. Likewise, Piantadosi et al. (2012) argued that lexical ambiguity is an efficient property of any communication system when communication is contextualized.

@@

But at what level of does local, in-the-moment interaction lead to optimized structure at the level of the lexicon? The example given by Zipf (1935) appears to describe local interaction in terms of a \textit{reference game}. Speakers intend to refer to some object in the world $m$. They choose some utterance $u$ to transmit this intended meaning, $u \rightarrow m$. The listener attempts to reconstruct this intended meaning given the transmitted utterance $m \rightarrow u$. Other functionalist projects have also assumed this basic reference game setting (Regier, et al. (201X); Piantadosi et al. (2012); Kirby, Smith & Folds (201X), OTHERS) and this simplification of the communicative act has proven productive in both theoretical (Ferrer-i-cancho, et al.,  201X), simulation-based (iterated learning stuff) and empirical explorations of efficient language structure and use (Some of Robert’s work here).

This move to study efficient language use in terms of reference games has clear theoretical underpinnings as well. The linguist Lawrence Horn (1984), highlighted the importance of Zipf’s principles for explaining conversation-level phenomena. Horn suggested a link between Zipf’s conception of speaker- and listener-effort and ideas related to conversational pragmatics by Grice (1975), arguing that the interaction of Zipf’s forces were “largely responsible for generating Grice’s conversational maxims and the schema for pragmatic inference derived therefrom.” Put differently, the system-level efficiencies we see in language are fundamentally related to the local-level efficiencies that occur during rational, pragmatic conversation.\par

In this work, we present a step toward formalizing the framing presented above  -- connecting Zipfian notions of efficient language structure to Gricean notions of rational conversation. To do so minimally requires three basic ingredients -- (1) a language property we’d like to explain, (2) a framework for describing in-the-moment interactions of speaker-listeners, and (3) some measure of linguistic efficiency. For the current project we focus on the “communicative function of ambiguity” for our property -- translating the functionalist theory pursued by Piantadosi et al. (2012) into our framework. To model local, conversational interactions (2) we adopt the Rational Speech-act framework (RSA) (Frank & Goodman, 2012; Goodman & Frank, 2016). Finally, (3) we derive an objective function for measuring language efficiency in the reference game setting.\par

The current work examines the question, “how does having \textit{lexical ambiguity} make using language $\ell$ more or less useful for communication”? We begin with a high-level introduction to the modelling framework introducing the basic ingredients we will need to represent language as repeated reference games. Following this introduction we derive a simple objective function for measuring the efficiency of linguistic systems in this setting. We move on to two case-studies examining the questions posed above, examining the efficiency of ambiguity from a language design and language use perspective.

# Exploring efficient language- design and use in rational pragmatic agents

```{r plot-reference-game, fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=2.5, fig.height=2.5, set.cap.width=T, num.cols.cap=1, fig.cap = "An example reference game with associated literal semantics (in our terminology a ‘language’)."}
img <- png::readPNG("figs/game1.png")
grid::grid.raster(img)
```

### Reference games
Zipf’s canonical (1935) example of optimal speaker- and listener-languages took the form of a reference game (similar to those described by Wittgenstein). We adopt that formulation here, assuming these communication games as our basic unit of analysis. In this setting, speakers and listeners are aware of a set of objects $M$, which we will refer to as *meanings* and are knowledgeable about the set of possible signals $U$ (*utterances*) that can be used to refer to a given meaning. Utterances may have different relative costs, operationalized via a prior over utterances $P(U)$. Similarly, meanings differ in the relative degree to which they need to be talked about, operationalized as a prior over meanings $P(M)$. Note that the prior over meanings are analogous to the *need probabilities* assumed in previous work (Regier, CITATION). We consider a set of contexts $C$ with an associated prior $P(C)$. Importantly, contexts describe different need probability states, formalized as different conditional distributions over meanings e.g. $p(M|C=c_i)$. Finally, we consider a set of communicative events $e \in E$ where $<u, m> = e$ is a tuple of utterance, referent pairs.

@@

describes the literal semantics of $\ell$. E.g. the language describes semantic mappings $[\![u_1]\!]\ = \{m_1, m_2\}, [\![u_2]\!]\ = \{m_2\}, [\![u_3]\!]\ = \{m_3\}$.

### Speakers and listeners
The Rational Speech-act framework (RSA) is computational-level theory of pragmatic language use. RSA can largely be understood as a formalization of essential Gricean pragmatic principles -- interlocutors reason about one another and their shared context. For this reason, we adopt RSA as our representational framework to operationalize Gricean (rational and pragmatic) speaker-listeners. For a more extensive description of RSA as well as its applications we refer readers to Goodman & Frank (2016), limiting the current discussion to essential, high-level details of the framework. \par

An RSA *speaker agent* defines a conditional distribution over utterances, mapping from intended meanings $M$ to utterances $U$ using $\ell$. That is, a speaker defines $P_{speaker}(u|m;\ell)$. We will use $S(u|m;\ell)$ as short-hand throughout.  A *listener agent* defines a conditional distribution over meanings, mapping from utterances $U$ to meanings $M$ using $\ell$. We will use $L(m|u; \ell)$ as shorthand. Note that both speakers and listeners can induce joint distributions over the set of all signaling events $E$, although, importantly, these distributions may differ:
$$P_{speaker}(u, m; \ell) = S(u|m; \ell)p(m)$$
$$P_{listener}(u, m; \ell) = L(m|u; \ell)p(u)$$

In general, we would like to consider the efficiency of a system $\ell$ in terms of these joint distributions.

@@

Zipf proposed that the particular distributional properties found in natural language emerge as a result of competing speaker and listener pressures. We operationalize this in equation (1) -- the efficiency of a linguistic system $\ell$ being used by speaker and listener agents $S$ and $L$ is the sum of the expected speaker and listener effort to communicate over all possible communicative events $E$.

\begin{equation}
\text{Efficiency}(S, L, \ell) = \mathbb{E}_{e \in E}[\text{speaker effort}] + \\ \mathbb{E}_{e \in E}[\text{listener effort}]
\end{equation}

Let speaker effort be the negative log probability (surprisal) of a particular utterance. Intuitively, the number of bits needed to encode the utterance $u$.

$$\text{speaker effort} = -log_2(p(u))$$

Let listener effort be the negative log probability a listener disambiguates an intended meaning $m$ given an utterance $u$ Intuitively, the number of guesses a listener would need to discover the intended meaning $m$ given an utterance $u$.
$$\text{listener effort} = -log_2(L(m|u; \ell))$$

Rewriting (1) we have
\begin{equation}
\begin{split}
\text{Efficiency}(S, L, \ell) = \mathbb{E}_{e \in E}[-log_2(p(u))] + \\
\mathbb{E}_{e \in E}[-log_2(L(m|u; \ell))]
\end{split}
\end{equation}
In general, these expectations are each taken over the set of possible communicative events $e\in E$ weighted by their probability, $p(e)$. Recall this is the set of all utterance, meaning pairs $<u, m> = e \in E$.

\begin{equation}
 = \sum_{e \in E}p(e)[-log_2(p(u))] + \sum_{e \in E}p(e)[-log_2(L(m|u; \ell))]
\end{equation}

We assume that the particular joint distribution over utterance-meaning $<u, m> = e$ pairs follows from a simple generative model. First, some meaning is sampled with probability $p(m)$. Our speaker attempts to convey this intended meaning to a listener via an utterance $u$ by sampling from the speaker conditional distribution $S(u|m; \ell)$. Combining these terms leads to the *speaker's joint distribution over events* which we can write as $P(e) = S(u | m; \ell)p(m) = P_{speaker}(u, m; \ell)$.

\begin{equation}
\begin{split}
  = \sum_{u, m}P_{speaker}(u, m; \ell)[-log_2(p(u))] + \\ 
   \sum_{u, m}P_{speaker}(u, m; \ell)[-log_2(L(m|u; \ell))]
\end{split}
\end{equation}

Simplifying we arrive at (5):
\begin{equation}
\begin{split}
 = \sum_{u, m}P_{speaker}(u, m; \ell)[-log_2(L(m|u; \ell)p(u))]
 \end{split}
\end{equation}

Note that $L(m|u; \ell)p(u)$ is the listener-based joint distribution over all communicative events ($P_{listener}(u, m; \ell)$).
\begin{equation}
  = \sum_{u, m}P_{speaker}(u, m; \ell)[-log_2(P_{listener}(u, m; \ell))]
\end{equation}
This is the simply the cross-entropy between the speaker an listener joint distributions.
\begin{equation}
\begin{split}
  = \mathbb{E}_{P_{speaker}}[-log_2(P_{listener})] \\
  = H_{cross}(P_{speaker}, P_{listener})
\end{split}
\end{equation}
From an information-theoretic perspective this objective is intuitive. Cross-entropy gives us a measure of dissimilarity between two distributions -- the average number of bits required to communicate under one distribution, given that the “true” distribution differs. In our case, this is the difference between the joint distribution assumed by the speaker $P_{speaker}$ and listener $P_{listener}$. A good language $\ell$ used by a set of a pair of speaker-listeners will have properties which minimize this objective.\par

# Simulations

Ambiguity is ubiquitous in natural language. At a minimum, the listener task in everyday communication is marked by the frequent need to handle lexical ambiguity (words often have multiple meanings) as well as syntactic ambiguity (sentences often have multiple parses), in addition to other instances of ambiguity (Wasow, Perfors & Beaver, 201X). Chomsky (2002) famously claimed that the presence of ambiguity in natural language provides evidence that language has not been optimized for communication, stating “If you want to make sure that we never misunderstand one another, for that purpose language is not well designed, because you have such properties as ambiguity.” Note that this analysis is analogous to the Zipfian optimal Listener language $\ell_{listener}^*$ we described earlier and define in terms of equation (9).\par

Piantadosi et al. (2012) argue just the opposite. They claim that ambiguity is an *efficient* property of any communication system in which *communication is contextualized*. Simply put, it is useful to have a language that re-uses low-cost material (has ambiguity) so long as the cost of disambiguating the material is low. In particular, context (or common ground) can provide useful information for disambiguation. Say we have two objects in the world $m_1$ and $m_2$ and two languages $\ell_1$ and $\ell_2$. In language $\ell_1$ the low-cost utterance $u_1$ can be used to refer to both $m1$ and $m_2$ (i.e. $[[u_1]]_{\ell_1} = \{m_1, m_2 \}$. Language $\ell_2$, by contrast, defines a mapping such that the low cost utterance $[[u_1]]_{\ell_2} = \{m_1\}$ and a higher cost utterance $[[u_2]]_{\ell_2} = \{m_2 \}$. While it is cheaper for a speaker to use $\ell_1$ (because she can always use a low-cost utterance), it is more difficult for a listener (because $u_1$ is ambiguous). However, if the meaning $u_1$ can be disambiguated *in context* then the listener will not incur disambiguation cost and the speaker can always use the low-cost utterance. This a basic argument for the contextual efficiency of polysemy. If I use the word “bat” at a baseball game a listener will likely infer I am referring to the baseball sense of the word. If I use the word “bat” in a dark cave, a listener will likely infer I am referring to the animal.

In the following experiments we explore two aspects of Piantadosi’s claim. First, we examine the efficient language *structure* aspect of their claim, exploring when the optimal linguistic system $\ell^*$ (under our cross-entropy objective) is most likely to contain ambiguous lexical items. In the second experiment we explore an efficient language *use* aspect of the question -- at what point in a conversation is it useful for a speaker to use ambiguous lexical material?

## Simulation 1:  Optimal languages contain ambiguity when context is informative

The argument outlined by Piantadosi et al. (2012) is compatible with our current reference game setting with just a single addition -- context. The authors argue that mapping two meanings $m_1$ and $m_2$ to a single utterance $u_1$ is useful when they can be disambiguated *in context*. In our case we consider a context $c$ to specify a unique ordering of the need probabilities. That is, $p(m|c_1) \neq p(m|c_2)$, when there are two contexts $|C| = \{c_1, c_2\}$.

This in turn, leads to an update to our linguistic efficiency objective as we now would like to consider the average speaker-listener effort over all contextualized communicative events. That is,

\begin{equation}
\begin{split}
\sum_{c\in C}p(c)\sum_{u, m}S(u|m, c;\ell)p(m|c)-log[L(m|u,c;\ell)p(u)] = \\ \mathbb{E}_{c\sim P(c)}[\mathbb{E}_{P_{speaker}(u, m, c; \ell)}[-log(P_{listener}(u, m, c; \ell))]]
\end{split}
\end{equation}

Note that in the case that $|C| =1$, our objective simplifies to our original equation (7). With this update we can compare languages $L$ while varying the degree to which context contains useful information, varying the number of contexts (the size of $|C|$).\par

### Simulation set-up

#### Basic ingredients
We conduct $N=2000$ simulations. For each simulation we enumerate the set of *valid* languages in which $|U|=|M|=4$. Note that a language $\ell \in L$ is "valid" so long as each possible meaning in $m \in M$ can be referred to by at least one form $u \in U$ (every column of $\ell$ has some non-zero assignment). For a given simulation the goal is to find the language $\ell^*$ which minimizers our cross-entropy objective.\par

Recall that language efficiency is both a function of the particular semantic mappings induced by that language, the speaker and listener agents ($S$ and $L$), as well as the utterance ($P(U)$), meaning ($P(M)$), and context priors ($P(C)$). Rather than assume particular structure for our utterance, meaning and context prior distributions, for each simulation we generate $P(U) \sim \text{Dir}(1, |U|)$, $P(M) \sim \text{Dir}(1, |M|)$, and $P(C) \sim \text{Dir}(1, |C|)$ where $\text{Dir}(1, k)$ specifies the uniform Dirichlet distribution over a $k$-dimensional probability vector.

#### Context
Following the argument given by Piantadosi et al. (2012) we want to assess the impact of *context* on our objective. To do so we consider four conditions with $n=500$ simulations each (that is, 500 unique sets of $\{P(U), P(M), P(C)\}$. Our first is a *single-context* condition ( $|C|=1$) -- there is a only a single context describing $P(M)$. Our second condition contains two-contexts  ($|C| = 2$) -- we consider efficiency under both $P(M|C=c_1)$ as well as $P(M|C=c_2)$. The third and fourth condition correspond accordingly with $|C|=3$ and $|C| = 4$, respectively.

#### Baselines
For comparison, we also examine properties of optimal languages under two additional objectives. Zipf (1935) proposed that the optimal speaker language $\ell_{speaker}^*$ should only optimize speaker effort. We operationalize this using the *first half* of equation (1).

\begin{equation}
\ell_{speaker}^* = argmin_{\ell\in L}\mathbb{E}_{P_{speaker}(u, m; \ell)}(p(u))
\end{equation}

The optimal listener language $\ell_{listener}^*$, by contrast, should only optimize listener effort. We operationalize this using the *second half* of equation (1).

\begin{equation}
\ell_{listener}^* = argmin_{\ell\in L}\mathbb{E}_{P_{speaker}(u, m; \ell)}(L(m|u;\ell))
\end{equation}

#### Hypotheses
Piantadosi’s claim leads to the following set of hypotheses: (1) The probability that an optimal language contains ambiguity under our Cross-Entropy objective should increase with the number of contexts (as $|C|\rightarrow\inf$). (2) The optimal speaker language under our “Speaker-only” objective should always map all meanings to the single, lowest cost utterance. (3) The optimal listener language should never contain ambiguous material.

### Results
Figure 1. plots the proportion of optimal languages under $H_{cross}(P_{Speaker}, P_{Listener}; \ell)$ for each context condition. The red line indicates that as the number of contexts increases, so does the probability that the optimal language $\ell^*_{cross}$ under our speaker-listener cross-entropy objective contains ambiguity. We also plot the proportion of optimal speaker-only $\ell^*_{speaker}$ and listener-only $\ell^*_{listener}$ languages that contain ambiguity. In line, with Zipf's predictions, if languages are designed only to minimize speaker effort then optimal languages will assign all meanings to a single utterance (full ambiguous). Likewise, if languages are designed only to minimize speaker listener effort then ambiguity should always be avoided (bijective form-meaning mappings).\par

```{r plot-optimal-langs, fig.env = "figure*", fig.pos = "h", fig.width=8, fig.height=2, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Optimal languages are more likely to contain ambiguous items as amount of contextual information increases. Vertical axis shows the proportion of optimal languages that contain ambiguity. Horizontal axis shows the number of contexts in each condition (1-4). Red-line represents the optimal language under our Zipfian cross-entropy objective while the blue and red lines show optimal languages under speaker- and listener-only consideration."}

@@

### Summary
Piantadosi et al. (2012) argued that it is useful to re-use low-cost linguistic forms for multiple meanings when they can be disambiguated in context. We simulated this effect in a reference game setting, examining the probability that an optimal language will contain an word with an ambiguous mapping as a function of context size. Using our speaker-listener cross-entropy measure of efficiency we showed that optimal languages are more likely to have ambiguous items when context is informative, but not under objectives which emphasizes speaker- or listener-effort alone.\par

## Simulation 2: Rational-pragmatic speakers use ambiguity efficiently

Our first experiment made a fairly direct test of the communicative function of ambiguity theory presented by Piantadosi et al. (2012). Sampling a set of need probabilities and utterance costs, we explored the space of possible languages, examining the properties of the language $\ell^*$ which minimized our objective. Results indicated that ambiguity is an efficient property when context is informative. This simulation, however, assumed that our speaker and listener agents could always disambiguate items context. That is, in our four conditions both agents had perfect knowledge of the conditional distributions $P(M|C_i)$.\par

This is a strong assumption and likely unrealistic for much of day-to-day communication.  In talking with another person, we often \textit{don’t have perfect knowledge of the current context of a conversation}. We can consider this notion of “context” as analogous to a *topic* of conversation. Consider the scenario in which a reader is beginning a newspaper article. While they may have some knowledge about the article’s topic (perhaps from the title), they may not have complete knowledge of its contents, including the persons or events involved. In this scenario, using an ambiguous pronoun early in the article lead to misunderstanding as there isn’t sufficient context for disambiguation. Using such ambiguity is only warranted when the appropriate person or event is salient in the context. \par

We consider such a scenario in simulation two -- analyzing efficient *use* of ambiguity over discourse. In this scenario, we assume a single language $\ell$ which contains both ambiguous and unambiguous utterances. The ambiguous utterances are less costly than the unambiguous utterances. Crucially, we do not assume that a listener has access the conditional distributions $p(M|C=c)$ a priori. Rather, we assume that the listener has knowledge of the set of possible contexts $C = \{c_1, \dots, c_k\}$, but \textit{does not know which one is currently being used by the speaker}. In this setting, a speaker minimizing her production effort would like to use the low-cost material as frequently as possible. However, this will incur a disambiguation cost to the listener if they are uncertain of the current context $c_{current}$ (topic of conversation).\par

Over the course of a discourse, $D$ the listener tries to infer both the current topic, $c_{current}$, as well as the particular meaning $m$ of a given utterance $u$. That is we consider a listener model $L(m, c|u,D;\ell)$ and speaker model $S(u|m,c,D;\ell)$. Note that both the speaker and listener can track the history of previous utterances $D$. Importantly, a listener can attempt to infer the current topic of conversation $c_{current}$ using the discourse history $D$.\par

```{r plot-optimal-use, fig.env = "figure*", fig.pos = "h", fig.width=8, fig.height=2, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Left facet (A) shows the empirical probability that our speaker used an ambiguous utterance as a function of discourse position. As the discourse proceeds, the speaker is more likely to use an ambiguous term. Right face (B) shows the probability the listener assigns to the current topic of conversation."}
img <- png::readPNG("figs/fig2.png")
grid::grid.raster(img)
```

### Hypothesis
In this setting, if speakers are *using ambiguity efficiently* they should only use the ambiguous material when they know the listener can disambiguate the items in context. Since listeners accumulate information about the current topic distribution ($P(C=c_{current}|D)$) over as the discourse progresses, we should expect to see more frequent use of ambiguous items later in the discourse.

### Basic ingredients

We conduct $N=200$ simulations, generating discourses of length $|D|=30$ utterances.  We consider a single language $\ell$ with $|U|=6$ and $|M|=4$ specified by the boolean matrix:

>>>>

Why do languages look the way they do? Zipf [-@Zipf1949a] proposed that distributional properties found in natural language were evidence of speaker-listener effort minimization. In his own words, “we are arguing that people do in fact act with a maximum economy of effort, and that therefore in the process of speaking-listening they will automatically minimize the expenditure of effort.” Evidence for this claim was largely derived at the level of the lexicon. Zipf argued that the particular relationship between a word’s frequency and its rank, length, and denotation size could be explained as an emergent property of speaker-listener effort minimization. \par

Zipf articulated what is now considered a *functionalist* approach to language science -- analyzing language structure and use in terms of efficiency. Such an approach might reframe our opening question as follows -- how does having \textbf{property x} make using language $\ell$ more or less useful for communication? This reframing provides an opportunity to study both language *structure* (as Zipf primarily did) or *use* in terms of efficiency. For example, Regier et al. [-@RegierKempKay2015a] showed that languages appear to organize semantic domains (form-meaning mappings) in maximally efficient ways. Likewise, Piantadosi et al. [-@Piantadosi2011a] argued that lexical ambiguity is an efficient property of any communication system when communication is contextualized.\par

@@

But how does such optimization take place? The example given by Zipf [-@Zipf1949a], he appears to describe local, communicative interactions in terms of a \textit{reference game}. Speakers intend to refer to some object in the world $m$. They choose some utterance $u$ to transmit this intended meaning, $u \rightarrow m$. The listener attempts to reconstruct this intended meaning given the transmitted utterance, $m \rightarrow u$. Other functionalist projects have also assumed this basic reference game setting [@RegierKempKay2015a; @Piantadosi2011a] and this simplification of the communicative act has proven productive in theoretical [@FerreriCancho2018a], simulation-based [@KirbyGriffithsSmith2014a] and empirical explorations [@HawkinsFrankeSmithGoodman2018a] of efficient language structure and use.\par

This move to study efficient language use in terms of reference games has clear theoretical underpinnings as well. The linguist Lawrence Horn highlighted the importance of Zipf’s principles for explaining conversation-level phenomena. Horn suggested a link between Zipf’s conception of speaker- and listener-effort and ideas related to conversational pragmatics by the philosopher Paul Grice, arguing that the interaction of Zipf’s forces were “largely responsible for generating Grice’s conversational maxims and the schema for pragmatic inference derived therefrom” [@Horn1984a]. Put differently, the system-level efficiencies we see in language are fundamentally related to the local-level efficiencies that occur during rational, pragmatic conversation.\par

In this work, we present a step toward formalizing the framing presented above  -- connecting Zipfian notions of efficient language structure to Gricean notions of rational conversation [@Grice1975a]. To do so minimally requires three basic ingredients -- (1) a language property we’d like to explain, (2) a framework for describing in-the-moment interactions of speaker-listeners, and (3) some measure of linguistic efficiency. For the current project we focus on the “communicative function of ambiguity” for our property -- translating the functionalist theory pursued by Piantadosi, et al. [-@Piantadosi2011a] into our framework. To model local, conversational interactions (2) we adopt the Rational Speech-act framework (RSA) [@FrankGoodman2012a; @GoodmanFrank2016a]. Finally, (3) we derive an objective function for measuring language efficiency in the reference game setting.\par

The current work examines the question, “how does having \textit{lexical ambiguity} make using language $\ell$ more or less useful for communication”? We begin with a high-level introduction to the modelling framework introducing the basic ingredients we will need to represent language as repeated reference games. Following this introduction we derive a simple objective function for measuring the efficiency of linguistic systems in this setting. We move on to two case-studies exploring the questions posed above -- examining the efficiency of ambiguity from a language design and language use perspective.\par

# Exploring efficient language design and use in rational pragmatic agents

```{r plot-reference-game, fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=2.5, fig.height=2.5, set.cap.width=T, num.cols.cap=1, fig.cap = "An example reference game with associated literal semantics (in our terminology a ``language'')."}
img <- png::readPNG("figs/game1.png")
grid::grid.raster(img)
```

### Reference games
Zipf’s example of optimal speaker- and listener-languages took the form of a reference game (similar to those described by Wittgenstein). We adopt that formulation here, assuming these communication games as our basic unit of analysis. In this setting, speakers and listeners are aware of a set of objects $M$, which we will refer to as *meanings* and are knowledgeable about the set of possible signals $U$ (*utterances*) that can be used to refer to a given meaning (see Figure 1 for an example reference game). Utterances may have different relative costs, operationalized via a prior over utterances $P(U)$. Similarly, meanings differ in the relative degree to which they need to be talked about, operationalized as a prior over meanings $P(M)$. Note that the prior over meanings are analogous to the *need probabilities* assumed in previous work Regier, Kemp & Kay [-@RegierKempKay2015a]. We consider a set of contexts $C$ with an associated prior $P(C)$. Contexts describe different orderings of the underlying meaning prior, formalized as different conditional distributions over meanings e.g. $p(M|C=c_i)$. Finally, we consider a set of communicative events $e \in E$ where $<u, m> = e$ is a tuple of utterance, meaning pairs.\par

@@

describes the literal semantics of $\ell$.

### Speakers and listeners
The Rational Speech-act framework (RSA) is computational-level theory of pragmatic language use. RSA can largely be understood as a formalization of essential Gricean pragmatic principles -- interlocutors reason about one another and their shared context. For this reason, we adopt RSA as our representational framework to operationalize Gricean (rational and pragmatic) speaker-listeners [@Grice1975a]. We refer the reader to Section 1 of our supplementary materials for a more detailed description of basic RSA speaker-listener definitions, and to Goodman & Frank [-@GoodmanFrank2016a] for an overview of its appliations. We limit the current discussion to essential, high-level details of the framework.\par

An RSA *speaker agent* defines a conditional distribution over utterances, mapping from intended meanings $M$ to utterances $U$ using $\ell$. That is, a speaker defines $P_{speaker}(u|m;\ell)$. We will use $S(u|m;\ell)$ as short-hand throughout.  A *listener agent* defines a conditional distribution over meanings, mapping from utterances $U$ to meanings $M$ using $\ell$. We will use $L(m|u; \ell)$ as shorthand. Note that both speakers and listeners can induce joint distributions over the set of all events $E$, although, importantly, these distributions may differ:
$$P_{speaker}(u, m; \ell) = S(u|m; \ell)p(m)$$
$$P_{listener}(u, m; \ell) = L(m|u; \ell)p(u)$$

@@

Zipf (1949) proposed that the particular distributional properties found in natural language emerge from competing speaker and listener pressures. We operationalize this in equation (1) -- the efficiency of a linguistic system $\ell$ being used by speaker and listener agents $S$ and $L$ is the sum of the expected speaker and listener effort to communicate over all possible communicative events $E$.\par

\begin{equation}
  \text{Efficiency}(S, L, \ell) = \mathbb{E}_{e \sim P(E)}[\text{speaker effort}] + \mathbb{E}_{e \sim P(E)}[\text{listener effort}]
\end{equation}

We assume that speaker effort is the negative log probability of an utterance $-log_2(p(U))$, listener effort is the negative log probability of the listener's conditional $-log_2(L(M|U))$, and that events $e = <u, m>$ are sampled according the to following generative model -- some object occurs in the world with probability $p(M=m)$, the speaker attempts to refer to that object by sampling from her conditional distribution $S(u|m)$ (i.e. $e \sim p(m)S(u|m)$). Using these basic ingredients we arrive at the following objective:

\begin{equation}
\begin{split}
  = \sum_{u, m}P_{speaker}(u, m; \ell)[-log_2(P_{listener}(u, m; \ell))]
\end{split}
\end{equation}
This is the simply the cross-entropy between the speaker and listener joint distributions.
\begin{equation}
\begin{split}
  = \mathbb{E}_{P_{speaker}}[-log_2(P_{listener})] \\
  = H_{cross}(P_{speaker}, P_{listener})
\end{split}
\end{equation}
From an information-theoretic perspective this objective is intuitive. Cross-entropy gives us a measure of dissimilarity between two distributions -- the average number of bits required to communicate under one distribution, given that the “true” distribution differs. In our case, this is the difference between the joint distribution assumed by the speaker $P_{speaker}$ and listener $P_{listener}$. An "efficient" language $\ell$ used by a set of a pair of speaker-listeners will have properties which minimize this objective. For a complete derivation and its componetns we refer the reader to Section 2.1 of our supplementary materials. \par

# Simulating the communicative function of ambiguity

Ambiguity is ubiquitous in natural language. The task of understanding language is marked by a frequent need to handle lexical ambiguity (words often have multiple meanings) as well as syntactic ambiguity (sentences often have multiple parses), in addition to many other instances of ambiguity [@WasowPerforsBeaver2005a]. Chomsky [-@Chomsky2002a] famously claimed that the presence of ambiguity in natural language provides evidence that language has not been optimized for communication, stating “If you want to make sure that we never misunderstand one another, for that purpose language is not well designed, because you have such properties as ambiguity.” Note that this analysis is analogous to the Zipfian optimal Listener language $\ell_{listener}^*$ we described earlier and define in terms of equation 2 in section 2.2 for our supplementary materials.\par

Piantadosi et al. [-@Piantadosi2011a] argue just the opposite, claiming that ambiguity is an *efficient* property of any communication system in which *communication is contextualized*. Simply put, it is useful to have a language that re-uses low-cost material (has ambiguity) so long as the cost of disambiguating the material is low. In particular, context (or common ground) can provide useful information for disambiguation.\par

We provide a toy example of their argument here. Say we have two objects in the world $m_1$ and $m_2$, and two languages $\ell_1$ and $\ell_2$. In language $\ell_1$, the low-cost utterance $u_1$ can be used to refer to both $m1$ and $m_2$ (i.e. $[[u_1]]_{\ell_1} = \{m_1, m_2 \}$). With language $\ell_2$, however, the low cost utterance $u_1$ can only refer to $m_1$ and a higher-cost utterance $u_2$ can refer to $m_2$ (i.e. $[[u_1]]_{\ell_2} = \{m_1\}$ and $[[u_2]]_{\ell_2} = \{m_2 \}$). While it is cheaper for a speaker to use $\ell_1$ (because she can always use a low-cost utterance), it is more difficult for a listener (because $u_1$ is ambiguous). If context isn't disambiguating then we might prefer $\ell_2$ to $\ell_1$. But, if context is disambiguating then the speaker can use $u_1$ to refer to either $m_1$ or $m_2$ and $\ell_1$ should be preferred.\par

In the following experiments we explore two aspects of Piantadosi’s claim. First, we examine the efficient language *structure* aspect of their claim, exploring when the optimal linguistic system $\ell^*$ is most likely to contain ambiguous lexical items. In the second experiment, we explore an efficient language *use* aspect of the claim -- at what point in a conversation is it useful for a speaker to use ambiguous lexical material?\par

```{r plot-optimal-langs, fig.env = "figure*", fig.pos = "h", fig.width=8, fig.height=2, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Optimal languages are more likely to contain ambiguous items as the amount of contextual information increases. Vertical axis shows the proportion of optimal languages that contain ambiguity. Horizontal axis shows the number of contexts in each condition (1-4). Red-line represents the optimal language under our Zipfian cross-entropy objective while the blue and red lines show optimal languages under speaker- and listener-only consideration."}

@@

# Simulation 1:  Optimal languages contain ambiguity when context is informative

The argument outlined by Piantados et al. [-@Piantadosi2011a] is compatible with our current reference game setting with an addition -- context. The authors argue that mapping two meanings $m_1$ and $m_2$ to a single utterance $u_1$ is useful when they can be disambiguated *in context*. In our case we consider a context $c_i \in C$ to specify a unique ordering of the need probabilities. That is, $p(m|c_1) \neq p(m|c_2)$, when there are two contexts $|C| = \{c_1, c_2\}$.\par

This in turn, leads to an update to our linguistic efficiency objective as we now would like to consider the average speaker-listener effort over all contextualized communicative events (section 3 of supplementary materials).\par

## Simulation set-up

We conduct $N=2000$ simulations. For each simulation we enumerate the set of *valid* languages in which $|U|=|M|=4$. Note that a language $\ell \in L$ is "valid" so long as each possible meaning in $m \in M$ can be referred to by at least one form $u \in U$ (every column of $\ell$ has some non-zero assignment) and each form maps to at least one meaning (every column has some non-zero assignment). For a given simulation the goal is to find the language $\ell^*$ which minimizes our cross-entropy objective.\par

Recall that language efficiency is both a function of the particular semantic mappings induced by that language, the speaker and listener agents ($S$ and $L$), as well as the utterance ($P(U)$), meaning ($P(M)$), and context priors ($P(C)$). Rather than assume particular structure for our utterance, meaning, and context prior distributions, for each simulation we generate $P(U) \sim \text{Dir}(1, |U|)$, $P(M) \sim \text{Dir}(1, |M|)$, and $P(C) \sim \text{Dir}(1, |C|)$ where $\text{Dir}(1, k)$ specifies the uniform Dirichlet distribution over a $k$-dimensional probability vector.\par

### Context

Following the argument given by Piantadosi et al. [-@Piantadosi2011a], we want to assess the impact of *context* on our objective. To do so we consider four conditions with $n=500$ simulations each (that is, 500 unique sets of $\{P(U), P(M), P(C)\}$. Our first is a *single-context* condition ($|C|=1$) -- there is a only a single context describing $P(M)$. Our second condition contains two-contexts  ($|C| = 2$) -- we consider efficiency under both $P(M|C=c_1)$ as well as $P(M|C=c_2)$. The third and fourth condition correspond accordingly with $|C|=3$ and $|C| = 4$, respectively.\par

### Baselines

For comparison, we also examine properties of optimal languages under two additional objectives. Zipf [-@Zipf1949a] proposed that the optimal speaker language $\ell_{speaker}^*$ would only optimize speaker effort and the optimal listener language $\ell_{listener}^*$ would only optimize listener effort. We define these objectives using the first and second half of equation 1 and with greater detail in the supplementary materials (Section 2.2).\par

## Hypotheses

Piantadosi’s claim leads to the following set of hypotheses: (1) The probability that an optimal language contains ambiguity under our Cross-Entropy objective should increase with the number of contexts (as $|C|\rightarrow\inf$). (2) The optimal speaker language under our “Speaker-only” objective should always contain ambiguity (maximally re-using the lowest-cost utterance). (3) The optimal listener language should never contain ambiguous material.

## Results

Figure 1. plots the proportion of optimal languages under each objective, as a function of context condition. The red line shows that as the number of contexts increases, so does the probability that the optimal language $\ell^*_{cross}$ under our speaker-listener cross-entropy objective contains ambiguity. We also plot the proportion of optimal speaker-only $\ell^*_{speaker}$ and listener-only $\ell^*_{listener}$ languages that contain ambiguity. In line with Zipf's predictions, if languages are designed only to minimize speaker effort then optimal languages will assign lower-cost forms multiple meanings (all of the speaker-optimal languages contained ambiguity). Likewise, if languages are designed only to minimize listener effort then ambiguity should always be avoided (none of the listener-optimal languages contained ambiguity).\par

## Summary
